---
title: | 
    | **Spatial Exploratory Data Analysis: Enviornmental Justice Index**
subtitle: |
    | Thesis for a Master of Public Health, Epidemiology
author: |
    | Nathan Garcia-Diaz
    |
    | Brown University, School of Public Health
date: |
    |
    | `r format(Sys.Date(), '%B %d, %Y')`
mainfont: Times New Roman
fontsize: 11 pt
output:
  pdf_document:
  latex_engine: luatex
urlcolor: blue
include-before:
- '`\newpage{}`{=latex}'
---

# Purpose Statement 

This fill acts as an extension to the `Exploratory Data Analysis: Enviornmental Justice Index` by examining the spatial components. Given the nature of the data, an examination of the spatial components is required. Specifically, this document will pull from two sources to help with the writing of the code: Ch 7.6 and 7.7 in Analyzing US Census Data: Methods, Maps, and Models in R ([link](https://walker-data.com/census-r/index.html)) and Manny Gimond's A basic introduction to Moran’s I analysis in R ([link](https://mgimond.github.io/simple_moransI_example/#step_1:_define_neighboring_polygons)). 

The examination of the spatial components include: Moran's I calculation with Monte Carlos Simulation, Moran's Scatterplot, Local Spatial Autocorrelation with Getis-Ord local $G^*_i$, and Hot/Cold Spot Identification. 

Definitions:

* **Moran's I**: Moran's I is a measure of spatial autocorrelation, quantifying the degree to which a variable is similarly distributed across neighboring geographic areas. It ranges from -1 (indicating perfect dispersion) to +1 (indicating perfect clustering), with values around 0 suggesting a random spatial pattern. It is used to detect and measure the presence of spatial autocorrelation, helping analysts understand whether the spatial distribution of a variable is clustered, dispersed, or random.
* **Monte Carlo Simulations**: Monte Carlo simulation is a computational technique that uses repeated random sampling to estimate the statistical properties of a system. It is used in tandem with Moran's I calculation to assess the significance of observed spatial autocorrelation by comparing it to the distribution of Moran's I values generated under the null hypothesis of spatial randomness. This is preformed as suggested by Gimond.
* **Moran's Scatter Plot**: Moran's scatterplot is a graphical representation that illustrates the relationship between a variable's values and the spatially lagged values of the same variable, used to visualize spatial autocorrelation. The plot typically includes a 45-degree reference line and divides the data points into four quadrants to help identify patterns of clustering or dispersion. It is used to diagnose and visualize spatial autocorrelation, helping to identify patterns of spatial clustering or dispersion in a dataset.
* **Local Spatial Autocorrelation**: Local measures of spatial autocorrelation, like the Getis-Ord local $G^*_i$, are used to identify clusters or "hot spots" of similar values within a spatial dataset. The Getis-Ord local $G^*_i$ statistic specifically measures the degree of clustering of high or low values around each point, indicating areas with significant local spatial association.
    * *Positive Gi Values*: indicate areas where high values of rpl_themes are surrounded by other high values, or low values are surrounded by other low values. This suggests clustering of similar values.
    * *Negative Gi Values*: Indicate areas where high values of rpl_themes are surrounded by low values, or vice versa. This suggests spatial outliers or contrast.

```{r, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE, echo = FALSE)
knitr::opts_chunk$set(fig.width=6, fig.height=4) 
options(tigris_use_cache = TRUE)
options(repos = c(CRAN = "https://cran.r-project.org"))
```

\newpage

```{r preparation}
### importing packages
# define desired packages 
library(tidyverse)  # general data manipulation
library(knitr)      # Rmarkdown interactions
library(here)       # define top level of project folder
                        # this allows for specification of where 
                        # things live in relation to the top level
# spatial tasks
library(tigris)     # obtain shp files 
library(spdep)      # exploratory spatial data analysis


### loading data 
eji_df = read_csv(here::here("01_Data", "eji_df.csv")) %>% 
  mutate(geoid = as.character(geoid)) %>% 
  select(-...1)

# obtaining SPH files for RI tracts
tracts = tracts(state = "RI", year = 2010, cb = TRUE)

# removing excess characters to allow for join  
tracts$GEO_ID = str_remove(tracts$GEO_ID, "^1400000US")

# joining data 
eji_map = inner_join(tracts, eji_df, by = c("GEO_ID" = "geoid"))

## define neighbors
nb_list = poly2nb(eji_map, queen=TRUE)
```

\newpage 

```{r}
## assign weights
weights = nb2listw(nb_list, style="W", zero.policy=TRUE)
```

\newpage

# Morans I Calculation and Scatter Plot

At $\alpha = 0.05$, all variables are statistically significant. Only one that had a statistic close to zero,the estimated number of persons in group quarters (`e_groupq`). All other variables had higher Moran's I values that what was foud in the SVI. The variables with the top 3 highest moran's i value include the annual mean days above PM2.5 regulatory standard (3 yr avg) (`e_pm`), the probability of Contracting Cancer (assuming continuous exposure) (`e_totcr`), and the ambient concentrations of diesel (PM/m3) (`e_dslpm`). The summary index value remained higher (`rpl_eji`) than what was found for the SVI summary index value. All other statistically significant variables demonstrated weaker presence of spatial autocorrelation, and measures related to coal and mining did were statistically insignificant. 

Additionally, a moran's scatterplot was only preformed for the outcome of interest since creating graphs for all variables would provide redundant information. However, these graphs can be made available upon request. In support with the Moran's I calculation with Monte Carlo Simulations, the scatterplot suggests a positive correlation between the SVI Summary Value and its spatial lag, representative of spatial autocorrelation in the data.

```{r morans i and scatter plot}
## calculating test statistic - moran's i via monte carlo simulation
columns = as.data.frame(eji_map) %>% 
  select(rpl_eji, 
         starts_with("e_"),
         starts_with("ep_")) %>% 
  colnames()

moran_df = tibble(NULL)

## creates calculations for all variables
for (col in columns){
  # calculate the moran's object
  res = moran.mc(pull(eji_map, !!col), 
                 listw = weights,
                 nsim = 999, 
                 zero.policy = TRUE)
  
  # use the res object to create a new row in moran_df
  moran_df = moran_df %>% 
    bind_rows(tibble(variable = col, 
                     statistic = res$statistic, 
                     pvalue = res$p.value))
}

# creating moran scatter plot 
moran.plot(x = eji_map$rpl_eji, 
           listw = weights, 
           xlab = "EJI Summary Values", 
           ylab = "Neighbors Standardized EJI Summary Values",
           main = c("Moran Scatterplot for EJI Summary Values", 
                    "in Rhode Island, 2022"))
```

\newpage

```{r}
# creating summary table for moran's i values for all variables 
moran_df = moran_df %>% arrange(desc(statistic))
knitr::kable(moran_df, caption = c("Moran's I with Monte Carlo Simulations"), 
             digits = c(0, 3, 3))
```

\newpage 

# Local Spatial Autocorrelation with $G^*_i$ & Hot/Cold Spot Identification

```{r}
# For Gi*, re-compute the weights with `include.self()`
localg_weights = nb2listw(include.self(nb_list), style="W", zero.policy=TRUE)
eji_map = eji_map %>% 
  mutate(localG = as.numeric(localG(eji_map$rpl_eji, localg_weights)))

# Local spatial autocorrelation
ggplot(eji_map, aes(fill = localG)) + 
  geom_sf() +
  scale_fill_distiller(palette = "RdYlBu") + 
  theme_void() + 
  labs(fill = "Local Gi* statistic")
```

```{r}
# hotspot identification
eji_map = eji_map %>%
  mutate(Hot_Spot = case_when(
    localG >= 2.576 ~ "High cluster",
    localG <= -2.576 ~ "Low cluster",
    TRUE ~ "Not significant"
  ))

ggplot(eji_map, aes(fill = Hot_Spot), color = "grey90", size = 0.001) + 
  geom_sf() + 
  scale_fill_manual(values = c("red", "blue", "grey")) + 
  theme_void() + 
  labs(title = "Hot Spot for SVI Summary Values")
```

